---
title: "Dissecting Tweets from Donald Trump (and his alter-ego)"
author: "STAT 230"
date: "1/17/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

> http://varianceexplained.org/r/trump-tweets/


## Data

```{r}
library(twitteR)
library(lubridate)
library(wordcloud)
library(tidytext)
```


In order to obtain data from the Twitter API, you'll need to first setup a developer's account on Twitter via http://dev.twitter.com. 

```{r}
setup_twitter_oauth("consumer_key", "consumer_secret", "access_token", "access_secret")
trump_tweets <- userTimeline("realDonaldTrump", n = 3200)
trump_tweets_df_2 <- twListToDF(trump_tweets)
# saveRDS(trump_tweets_df_2, "trump_tweets_df_recent.rds")
# trump_tweets_df_2 <- readRDS("trump_tweets_df_recent.rds")
load(url("http://varianceexplained.org/files/trump_tweets_df.rda"))
trump_tweets_df <- rbind(trump_tweets_df_2, trump_tweets_df)
Encoding(trump_tweets_df$text) <- "latin1"
```

Data cleaning:

```{r}
tweets <- data.frame(text = trump_tweets_df$text,
                     date = with_tz(trump_tweets_df$created, "EST"),
                     source = trump_tweets_df$statusSource,
                     stringsAsFactors = FALSE)
table(tweets$source)
tweets$source <- gsub(".*Twitter for (.*)<.*", "\\1", tweets$source) 
table(tweets$source)

## select only the tweets from iphone or android
tweets <- tweets[tweets$source %in% c("iPhone", "Android"),]
```

Take a look at recent tweets from each device:

```{r}
head(tweets$text[tweets$source == "iPhone"])
head(tweets$text[tweets$source == "Android"])

```


## A few plots

```{r}
tweets_by_date <- aggregate(tweets$text~ date(tweets$date)+tweets$source,
                            FUN=length)
colnames(tweets_by_date) <- c("date", "source", "count")

plot(count ~ date, data = tweets_by_date, subset=source=="Android",
     type="l", lwd=2, col=2, xlab="Date",
     ylab="# Tweets by Device", ylim=c(0,35), xaxt="n")
xdates <- seq(from=as.Date("2015-12-01"), to=as.Date("2017-01-31"), by="months")
axis(1, labels=format(xdates, "%m/%y"), at=xdates )

lines(count ~ date, data = tweets_by_date, subset=source=="iPhone", lwd=2, col=3)
legend("topleft", col=2:3, lwd =2, c("Android", "iPhone"))
abline(v=as.Date("2016-11-08"), col="blue", lwd=2, lty=2)
```


```{r}
tweets_by_hour <- table(hour(tweets$date), tweets$source)
hrs <- as.numeric(rownames(tweets_by_hour))
plot(tweets_by_hour[,1]/sum(tweets_by_hour) ~ hrs, type="l", lwd=2, col=2, xlab="Hour",
     ylab="% of all Tweets by Device")
lines(tweets_by_hour[,2]/sum(tweets_by_hour) ~ hrs, lwd=2, col=3)
legend("topright", col=2:3, lwd =2, c("Android", "iPhone"))
```


```{r}
tweets_by_hour <- tweets_by_hour / rowSums(tweets_by_hour)
barplot(t(tweets_by_hour), xlab="Hour (EST)", 
        ylab="% of Hourly Tweets by Device", legend=colnames(tweets_by_hour))
```

## Examining Tweet Features

```{r}
tweets$quote <- "not quoted"
tweets$quote[grep('^"', tweets$text)] <- "double-quote"
tweets$quote[grep("^'", tweets$text)] <- "single-quote"

head(tweets$text[tweets$quote == "double-quote"])
head(tweets$text[tweets$quote == "single-quote"])

tweets$picture <- "no picture/link"
tweets$picture[grep("t.co", tweets$text)] <- "picture/link"
```

```{r}
tmp <- table(tweets$quote, tweets$source)
barplot(tmp, beside=TRUE,
         legend=rownames(tmp), ylab="# Tweets")

```

```{r}
barplot( table(tweets$picture, tweets$source), beside=TRUE,
         legend=c(unique(tweets$picture)), ylab="# Tweets")
```


## Tweets to Words

Some more cleaning:

```{r}
tweet_words <- tweets[tweets$quote == "not quoted",]
tweet_words$text <- gsub("https://t.co/[A-Za-z0-9]+", "", tweet_words$text)
tweet_words$text <- gsub("&amp;", "", tweet_words$text)

reg <- "([^A-Za-z\\d#@']|'(?![A-Za-z\\d#@]))"
tweet_words <- unnest_tokens(tweet_words, word, text, token = "regex", pattern = reg)
tweet_words <- tweet_words[grep("[a-z]", tweet_words$word),]
tweet_words <- tweet_words[!(tweet_words$word %in% stop_words$word),]
```


Comparing word frequencies

```{r}
word_freq <- sort(table(tweet_words$word), decreasing = TRUE)
word_freq_i <- sort(table(tweet_words$word[tweet_words$source=="iPhone"]),
                    decreasing = TRUE)
word_freq_a <- sort(table(tweet_words$word[tweet_words$source=="Android"]),
                    decreasing = TRUE)
```

Common words:

```{r}
par(las=1)
par(mar=c(5,8,4,2)) # increase y-axis margin.

barplot(word_freq[1:15], horiz=TRUE, main="Top 15 Tweeted Words", xlab="Count")
barplot(word_freq_i[1:15], horiz=TRUE, 
        main="Top 15 Tweeted Words (iPhone)", xlab="Count")
barplot(word_freq_a[1:15], horiz=TRUE, 
        main="Top 15 Tweeted Words (Android)", xlab="Count")
```

A better way to present this?

```{r}
wordcloud(names(word_freq), word_freq, min.freq=50, random.order=FALSE,
          scale=c(3, 0.5))
```


```{r}
# iphone
wordcloud(names(word_freq_i), word_freq_i, min.freq=30, random.order=FALSE,
          scale=c(3, 0.5))
# android
wordcloud(names(word_freq_a), word_freq_a, min.freq=30, random.order=FALSE,
          scale=c(3, 0.5))
```


To examining words that are more frequently used in one of the two sources than the other, we'll construct a ratio: 

$$AIratio = \log\left(\frac{\frac{\text{# in Android} +1}{\text{Total Android}+1}}{\frac{\text{# in iPhone}+1}{\text{Total iPhone}+1}}\right)$$

```{r, fig.height=8}
a_i_ratios <- table(tweet_words$word, tweet_words$source)
a_i_ratios <- a_i_ratios[rowSums(a_i_ratios) >= 10,]  
a_i_ratios <- log((a_i_ratios[,1]+1)/(sum(a_i_ratios[,1])+1))-
                                 log((a_i_ratios[,2]+1)/(sum(a_i_ratios[,2])+1))
a_i_ratios <- sort(a_i_ratios)

par(las=1)
par(mar=c(5,8,4,2)) # increase y-axis margin.
barplot(c(head(a_i_ratios, 10), tail(a_i_ratios, 10)), horiz=TRUE, 
        main="Frequently Tweeted Words", xlab="Android / iPhone log ratio")

```


## Sentiment Analysis

Details on lexicons:
> http://saifmohammad.com/WebPages/lexicons.html 

```{r}
nrc <- sentiments[sentiments$lexicon == "nrc", c("word", "sentiment")]
head(nrc)
table(nrc$sentiment)
```


```{r}
tweet_words_sent <- merge(tweet_words, nrc, by="word", all.x = TRUE)
head(tweet_words_sent)
```

Hmm... still some quotes... what's going on?

```{r}
tweets$text[tweets$date == "2016-10-22 21:26:32"] # doesn't work
class(tweets$date) # because date is being compared to character here

tweets$text[as.character(tweets$date) == "2016-10-22 21:26:32"] # attempt 2
tweets$quote[as.character(tweets$date) == "2016-10-22 21:26:32"] # attempt 2

```
Okay, so we haven't picked out all of the tweets with quotes. Could we fix this?


```{r}
grep("'", tweets$text) # finds indices of tweets with any single-quotes
grep("\\s+'", tweets$text) # finds indices of any occurrence of space followed by '
tweets$text[grep("\\s+'", tweets$text)]
```

```{r}
head(tweets$text[tweets$quote=="single-quote"])
```

Now, looking at the sentiments of the quotes more closely.

```{r}
sent_tbl <- table(tweet_words_sent$source, tweet_words_sent$sentiment)
mosaicplot(sent_tbl,
           main="Proportion of Sentiments Expressed by Source", las=1)
```

What are the frequently-used words that make up each sentiment?

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2, 2), las=1, mar=c(5,8,4,2)) # increase y-axis margin.
sents <- unique(nrc$sentiment)
sents <- sents[!(sents %in% c("negative", "positive"))]
for(s in sents) {
  tmp <- unique(tweet_words_sent$word[tweet_words_sent$sentiment == s])
  ratios <- a_i_ratios[names(a_i_ratios) %in% tmp]
  
  ratios <- head(ratios[order(abs(ratios), decreasing=TRUE)], 10)
  barplot(sort(ratios), horiz=TRUE, 
        main=s, xlab="Android / iPhone log ratio")  
}

```

