---
title: "The Titanic, Class Questionnaire (and Practicing R Best Practices)"
author: "STAT 230"
date: "January 24, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Best Practices, Tips & Tricks

Last class, we saw some of the basic ways that R can be used for exploratory 
data analysis. Today we'll spend some time getting a bit more acquainted with 
R, see some FAQ's you may encounter, and discuss some good habits to get into. 
We'll illustrate some of these on the raw class questionnaire dataset, 
of which a cleaned up version was provided for your homework assignment. 

* Configure RStudio options. Go to 'Tools' -> 'Global Options...'. There are a 
lot of options here you can select to help you form good coding habits. 

For example, I choose to add a vertical edge at 80 characters, so that I get into
the practice of not writing extremely long lines of code. I also select to 
indent via 2 spaces instead of a tab to save myself space. You're far less likely
to get text that trails off the page after compiling things if you stick to 
this rule.

* Set up organized folders on your computer for your work in this class (and
other classes in which you will be using R).

It can be extremely tempting to put all your R-related files into one place, such
as in the default **working directory**. This could cause all sorts of chaos as
the number of files increases throughout the semester.

* Maintain good coding style. 

We couldn't emphasize this enough. It isn't sufficient to write code that works 
for the moment being; you need to write code that you can read and interpret in
2 weeks, 4 weeks, etc. For that reason, following the following coding 
guidelines:

1. Always use `<-` for variable assignment. Ensure that all operators are 
preceded and followed by a space.

```{r}
a <- 5 # this is the right way to do it!

a<-5  # this seems harmless, does it?
a< -5 # how about this?
```


Here's an example of where it could all go wrong...

```{r}
a <- -5:5   # shorthand for seq(-5, 5, by=1)
a
a[a<-5]
a
```

For instances where you want to explicitly check whether, say, `a` is less than 
-5, put a space before and after the operator to remove the ambiguity.

```{r}
a < -5
```

2. Don't let yourself write lines of code longer than 80 characters. When you 
compile the file, say into Word, you might find that the larger font makes it
hard to have your code be wrapped around legibly. What happens if you need to 
write commands that are longer than 80 characters? Add returns and align 
appropriately. 

Here's what not to do:

```{r}
x <- data.frame(name=c("Anna", "Bob", "Casper", "Doris", "Ethan"), age=runif(5, min=0, max=50))
```

By the way, this is how we create a data frame. But really, it's better to 
format this code as follows:

```{r}
x <- data.frame(name=c("Anna", "Bob", "Casper", "Doris", "Ethan"), 
                age=runif(5, min=0, max=50))
```

You may have read from other places that it's possible to put multiple R commands
on the same line when separated by a `;`. Avoid doing this because it makes the
code less readable.

```{r}
x <- 5*exp(5); x  # there's no need to save yourself space. Just use a new line!
```

3. Indent your code.

As we get into topics like for-loops and if-statements, proper indentation will
significantly improve the readability of your code. (Thankfully, RStudio does
auto-indent for you, and a Cmd+I or Ctrl+I can help clean-up an existing mess.)


### Other Tips

* R is case sensitive. `x` is not the same thing as `X`.

* When the console is ready to take in a new command, you should see the line
begin with `>`. Sometimes it thinks you're still waiting to finish a previous
command, in which case, it shows `+`. Just click into the console, and hit
'Escape' on your keyboard and you should get back to the `>`.

* An R script (or Rmd script) runs sequentially from beginning to end. Do not
run code chunks out of order.

## Titanic Dataset

> Video clip: (1:10)
> https://www.youtube.com/watch?v=BQHaR_OdghI

This is a dataset that is stored as a .csv file, and is posted on Canvas. You should download this file and save it into an appropriate folder. We'll then read the file into R using the `read.csv()` function.

```{r, eval=F}
x <- read.csv("titanic.csv")
```

Did this work? If not, it's quite likely that your working directory is not the
one that contains the .csv file we're trying to load in. You'll have to either
put in the full path leading to this .csv file, or to set the proper **working
directory**. For example,

```{r}
titanic <- read.csv("titanic.csv")
colnames(titanic) <- tolower(colnames(titanic))
```

A good workflow when you start working with any dataset is to examine its dimensions and the first and last rows of the dataset.

```{r}
dim(titanic) # 1309 rows, 6 columns
head(titanic) 
tail(titanic)
```

You should only have to set the working directory once in a .Rmd file, generally. 
Keeping your data files organized in an appropriate directory structure will 
make this possible. It's also a good idea to set the working directory at the
beginning of your .Rmd file (say in the first code chunk).


The variable names above are generally obvious, with the exception of perhaps 
the following:

* `sibsp`: # of siblings or spouses aboard 
* `parch`: # of parents or children aboard

Here's a new command we haven't seen before that can help give us a sneak
preview of the data frame, its columns, and their types.

```{r}
str(titanic)
```

Each variable is summarized with its type ('num', 'int', or 'Factor'). 
'num' and 'int' are numeric or integer variables. Then we have 'Factor', which 
is actually R's way of encoding categorical variables. It's interesting to note
that factors look sort of like text values, but are actually represented as
numeric values on the back end (see the `str()` output above).

```{r}
head(titanic$pclass)
```

The distinct categories of a categorical variable can be seen using:

```{r}
levels(titanic$pclass)
```

The order in which these levels appear tells us what number gets assigned to 
each factor. (This detail is not particularly important for now, but will be
when we start looking at using categorical variables in linear regression.)


We can get a peek at some frequency tables:

```{r}
table(titanic$pclass)
table(titanic$survived)
table(titanic$sex)
```

We've long heard that women are more likely to have survived the Titanic disaster.
Do the data support this claim?

```{r}
table(titanic$sex, titanic$survived)
```

A graphical display may be more helpful here:

```{r}
# which one addresses the question better?
mosaicplot(sex ~ survived, data=titanic)
mosaicplot(survived ~ sex, data=titanic)

```

The first plot shows that a higher proportion of women survived compared to men.
The second plot shows that a higher proportion of survivors were women. 
To address the question of whether are more likely to survive, the first plot
does a better job of answering the question.

We can rotate the axis labels by tweaking the `las=` argument.

```{r}
mosaicplot(sex ~ survived, data=titanic, las=1)

```
We can make the axis text bigger by changing `cex=`.

```{r}
mosaicplot(sex ~ survived, data=titanic, cex=1.2)
```

Another way to do a mosaic plot is to use the `plot()` function again. 
Unfortunately, the `las=` axis label rotation functionality seems incompatible
with this approach.

```{r}
plot(sex ~ survived, data=titanic)
```

## Questionnaire Data

The New York City dataset we looked at last class is a nice first example because
it has already been scrubbed clean. Often, the data come to us in a relatively
messy format. For example, the class questionnaire....

```{r}
x <- read.csv("stat_230_survey_raw.csv")
```

** A bit of a sidebar:** Typically, we like to use the `as.is=TRUE` argument
when we run the `read.csv()` command. Here is a bit about why we might care. 
We'll start by checking the type of each column in the data frame:

```{r}
str(x)
```

Suppose we want to explore the hypothesis of whether the average pulse for 
all females is different than that for males. I threw the word **all** in there
to reflect that this is a question about the mean pulse in a population. We can
try to answer this question using our sample, of course, with an assumption that
our sample is a representative sample from the population.

Let's take a peek at the variables in question first to see what our sample
tells us.

```{r}
class(x$Gender)
table(x$Gender)
```
That stray 1 is actually a missing value for Gender, but is currently coded as a 
blank. Note that blanks appeared as one of the distinct values in the column in
the .csv file, so R thought that this is an appropriate level of the factor.

```{r}
levels(x$Gender)
```

The way that R is meant to encode missing values is to use an `NA`. 
Because we exported this data file from Google Docs, we have to recode 
this ourselves.

```{r}
x$Gender[x$Gender == ""] <- NA
levels(x$Gender)
table(x$Gender)
```

It's still there... how annoying. Lesson learned: factors can be annoying to 
clean up in R. You can't just discard a particular level or add one on the fly.
So to get around this problem, we can simply add `as.is=` in the `read.csv()`
command when we load in the data.

```{r}
x <- read.csv("stat_230_survey_raw.csv", as.is = TRUE)
x$Gender[x$Gender == ""] <- NA
table(x$Gender)
class(x$Gender)
```

So now, `Gender` is no longer a factor. `chr` stands for character, aka. text. That's okay, R will know to convert `Gender` to factor type when needed.


Let's examine the `Pulse` variable.

```{r}
summary(x$Pulse)
```


```{r}
boxplot(Pulse ~ Gender, data=x, ylab="Pulse")
```

Does it appear that the mean pulse for females might differ from the mean pulse
for males? Formally, we are asking about these two hypotheses:

$$H_0: \mu_{F} = \mu_M \;\;\;\;\;\text{ vs.  }\;\;\;\;\;H_a: \mu_F \ne \mu_M$$
where $\mu_F$ and $\mu_M$ are the mean pulses in the population of females and
males, respectively.

There are several ways we can answer this question.

### 2-Sample T-test

```{r}
t.test(Pulse ~ Gender, data=x)
```

How do we interpret the results here? We have a p-value of 0.8091. At any 
reasonable significance level $\alpha$ (say, 0.05), we do not have evidence to
suggest that the mean pulse differs between the genders. We also get a
95% confidence interval of (-6.16, 4.83). We might say that we are 95% confident
that the difference $\mu_F-\mu_M$ is between (-6.16, 4.83). How did we know it 
was $\mu_F-\mu_M$ and not $\mu_M-\mu_F$? Because in the line below 'sample 
estimates', we see 'Female' before 'Male'. 

Some questions you may have:

* The p-value here corresponds to a 2-sided alternative hypothesis. What can
I do if I wanted a 1-sided alternative hypothesis test instead? 

Questions like these can generally be answered by pulling up the help page, i.e.
`?t.test`. You will see that you can add the argument `alternative=` to be
'less' or 'greater', but that it defaults to 'two.sided'.

* What if I want a confidence interval for a different confidence level? 

The help file suggests that `conf.level=` is the argument to be modified.


