---
title: 'American Time Use Survey: Social/Leisure'
author: "STAT 230"
date: "2/23/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in the dataset:

```{r}
# x <- read.csv("atus_social.csv", as.is = TRUE)
x <- read.csv("https://www.dropbox.com/s/c7gidc2stjh3a2k/atus_social.csv?raw=1",
              as.is = TRUE)
head(x)
```

Some models we looked at:

```{r}
m1 <- lm(social_time ~ age, data=x)
summary(m1)

m2 <- lm(social_time ~ age + I(age^2), data=x)
summary(m2)
```


## Males vs. Females

```{r}
m3 <- lm(social_time ~ age + I(age^2) + sex, data=x)
summary(m3)
```

Predict the social time for a 20-year old male and for a 20-year old female.

```{r}
dat <- data.frame(age = c(20, 20), 
                  sex = c("M", "F"))
dat
predict(m3, dat)
```

Predict the social time for a 50-year old male and for a 50-year old female.

```{r}
dat <- data.frame(age = c(50, 50), 
                  sex = c("M", "F"))
dat
predict(m3, dat)
diff(predict(m3, dat))
```

Fitted equation:

$$\widehat{\text{social_time}} = 363 - 8.20 age + 0.119 age^2 + 43.0 sex_{Male}$$

$sex_{Male}$ is a vector of 0's and 1's. 0's where a row corresponds to a female and 1's where a row corresponds to a male. 

```{r}
head(model.matrix(m3))
```

Fitted equation for females:

$$\widehat{\text{social_time}} = 363 - 8.20 age + 0.119 age^2$$

Fitted equation for males:

$$\widehat{\text{social_time}} = 363 - 8.20 age + 0.119 age^2 + 43.0$$

$$\widehat{\text{social_time}} = 406 - 8.20 age + 0.119 age^2$$

```{r}
dat <- data.frame(age = 0:85,
                  sex = rep("M", 86))
p3 <- predict(m3, dat)

plot(social_time ~ age, data=x, pch=16, col=rgb(0, 0, 0, 0.1))
lines(dat$age, p3, col="blue", lwd=2)
```


Quick detour discussion on `rep()`:

```{r}
rep(1:3, each=2)
rep(1:3, times=2)
```


```{r}
dat <- data.frame(age = rep(0:85, times = 2),
                  sex = rep(c("M", "F"), each = 86))
p3 <- predict(m3, dat)

plot(social_time ~ age, data=x, pch=16, col=rgb(0, 0, 0, 0.1))
lines(dat$age[dat$sex == "M"], p3[dat$sex == "M"], col="blue", lwd=2)
lines(dat$age[dat$sex == "F"], p3[dat$sex == "F"], col="red", lwd=2)
```


```{r}
summary(m3)
```

Suppose we were to recode our model such that it fitted a coefficient for $sex_{Female}$, how would things change?

```{r}
class(x$sex)
levels(factor(x$sex))
```

The first level is "F" because of alphabetical order. In linear regression, the first level is taken to be the **baseline level**. We can force R to choose "M" to be the first level.

```{r}
x$sex <- factor(x$sex, levels=c("M", "F"))

m4 <- lm(social_time ~ age + I(age^2) + sex, data=x)
summary(m4)
```


The p-value for `sexF`. This is testing the hypotheses:

$$H_0: \beta_{sexF} = 0  \text{  vs.  }H_a:\beta_{sexF} \ne 0$$

If we fail to reject $H_0$, concluding that there is no evidence suggesting $\beta_{sexF}$ is nonzero, then that is like saying there is no difference in social time between the two sexes.

## Incorporating Diary Date

```{r}
head(x$diary_date)
class(x$diary_date)
```


A package called lubridate.

```{r}
#install.packages("lubridate")
library(lubridate)
x$diary_date <- ymd(x$diary_date)
head(x$diary_date)
class(x$diary_date)

x$wday <- wday(x$diary_date)
table(x$wday)


x$wday <- factor(wday(x$diary_date, label=TRUE), ordered=FALSE)
table(x$wday)
class(x$wday)
```

```{r}
m4 <- lm(social_time ~ age + I(age^2) + sex + wday, data=x)
summary(m4)
head(model.matrix(m4))
```


Let's consider modeling just a weekday vs. weekend effect, rather than breaking out exact days of the week.

```{r}
x$wknd <- x$wday %in% c("Sat", "Sun")
table(x$wknd, x$wday)
```

```{r}
m5 <- lm(social_time ~ age + I(age^2) + sex + wknd, data=x)
summary(m5)
head(model.matrix(m5))
```


Fitted equation:

$$\widehat{\text{social_time}} = 359 - 8.1 age + 0.119 age^2 - 42.6sex_{F} + 90.0 wknd$$


Males on a weekday:

$$\widehat{\text{social_time}} = 359 - 8.1 age + 0.119 age^2$$

Males on a weekend:

$$\widehat{\text{social_time}} = 449 - 8.1 age + 0.119 age^2$$

Females on a weekday:

$$\widehat{\text{social_time}} = 316.4 - 8.1 age + 0.119 age^2$$

Females on a weekend:

$$\widehat{\text{social_time}} = 406.4 - 8.1 age + 0.119 age^2$$