---
title: 'Class Questionnaire: Data Cleaning'
author: "STAT 230"
date: "Feb 2, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data Cleaning

Data cleaning is a critical first step in most data analysis. We will spend 
some time cleaning up the rest of this raw dataset. Let's start by loading in the
raw data again.

```{r}
x <- read.csv("stat_230_survey_raw.csv",
              as.is = TRUE)
str(x)
```

We will now iterate over each of the variables in turn to check for any issues 
that need fixing.

```{r}
summary(x$Pcredit)
hist(x$Pcredit)
```

```{r}
table(x$UorG)
```

There's a missing value here... see the 1 lingering on the left? In fact, let's
take a peek at this individual. 

```{r}
x[x$UorG == "",]
```

It appears that this person only answered the first question before submitting. 
How do we want to handle this row? Suppose we decide we simply don't trust the
data (not even the 1 for `Pcredit`). Then, we might decide to take this row out
entirely.

```{r}
x <- x[x$UorG != "",]
```


**A suggestion from last class...**

Can we just replace all blanks with NA simultaneously?

```{r}
x[x==""] <- NA
```

```{r}
table(x$UorG)
```

We can still abbreviate the labels:

```{r}
x$UorG[x$UorG == "Graduate student"] <- "Grad"
x$UorG[x$UorG == "Undergraduate student"] <- "Undergrad"
table(x$UorG)
```

Let's take a look at 'Year':

```{r}
table(x$Year)

x$Year[x$Year == "1 = freshman or first year grad"] <- "1"
unique(x$Year)
```

To save ourselves some typing:

```{r}
x$Year[grep("2 =", x$Year)] <- "2"
x$Year[grep("3 =", x$Year)] <- "3"
x$Year[grep("4 =", x$Year)] <- "4"
x$Year[grep("Beyond", x$Year)] <- "4+"
unique(x$Year)
table(x$Year)
```

```{r}
barplot(table(x$Year))
class(x$Year)
```

Height

```{r}
x$Height

# input: 'z' a vector that is supposedly numeric but has nonnumeric values
# output: a vector of indices corresponding to those nonnumeric values
findNonNumbers <- function(z) {
  return(which(is.na(as.numeric(z))))
}

temp <- findNonNumbers(x$Height)
temp
x$Height[temp[1]] # nothing to do, started as an NA
x$Height[temp[2]] <- 6*12+1
x$Height[temp[2]]

x$Height[temp[3]] <- 6*12+3
x$Height[temp[6]] <- 5*12+3
x$Height[temp[7]] <- 4*12+2
x$Height <- as.numeric(x$Height)
hist(x$Height)
```

Who are the outliers?

```{r}
x[which(x$Height > 100),]

x$Weight[which(x$Height > 100)] <- 110.23
x$Height[which(x$Height > 100)] <- 162/2.54

x[which(x$Height < 30),]

# if we believe these are all in units of feet, we can fix them together
x$Height[which(x$Height < 30)]
x$Height[which(x$Height < 30)] <- x$Height[which(x$Height < 30)]*12
hist(x$Height)
```

SAT scores:

```{r}
x$SAT_CR[findNonNumbers(x$SAT_CR)]
x$SAT_M[findNonNumbers(x$SAT_M)]
x$SAT_W[findNonNumbers(x$SAT_W)]

x[x == "n/a"] <- NA
x$SAT_CR <- as.numeric(x$SAT_CR)
x$SAT_M <- as.numeric(x$SAT_M)
x$SAT_W <- as.numeric(x$SAT_W)
```

```{r}
hist(x$High.school.GPA)
```

Television:

```{r}
x$Television[findNonNumbers(x$Television)]
x$Television <- as.numeric(x$Television)
hist(x$Television)

sort(x$Television)
```

```{r}
x[which(x$Television > 0 & x$Television < 5), 
  c("Television", "Web", "Exercise",
                                         "Academics")]
```



```{r}
temp <- findNonNumbers(x$Haircut)
x$Haircut[temp[2]] <- 10
x$Haircut[temp[5]] <- 45
x$Haircut[temp[6]] <- 0
x$Haircut <- as.numeric(x$Haircut)
hist(x$Haircut)
```

