---
title: 'American Time Use Survey: Social/Leisure'
author: "STAT 230"
date: "2/9/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Data

> https://www.bls.gov/tus/datafiles_2014.htm

We looked at this last class and got the data to a almost-clean enough format for us to work with. We quickly run the commands that get up to 
the point where we left off.


```{r}
load("atusresp.rda")
load("atussum.rda")
load("atuscps.rda")

social_cols <- names(atussum)
social_cols <- grep("t12", social_cols)
social_times <- rowSums(atussum[,social_cols])

x <- atussum[,c(1, 3:8)] # we like these columns!
x$social_time <- social_times
keep <- c("tucaseid", "num_children", "weekly_earn", "diary_date")
y <- atusresp[, keep]
x <- merge(x, y, by="tucaseid", all.x = TRUE)

keep <- c("tucaseid", "age", "sex", "famincome", "hh_size") # use age to check
y <- atuscps[atuscps$atus == 1,keep]
z <- merge(x, y,by="tucaseid", all.x=TRUE)

```

We were pondering this problem of strange differences in reported age
between ATUS and CPS for the same individuals. 

```{r}
table(z$age.x-z$age.y)
```

We decided that maybe we should toss out all of the individuals with a
discrepancy greater than 1 in absolute value. We did the following:

```{r}
# z <- z[abs(z$age.x-z$age.y) <= 1,]
```

Remove the -1's as well. 

```{r}
agediffs <- z$age.x - z$age.y
z <- z[agediffs == 0 | agediffs == 1,] 
z <- z[agediffs %in% 0:1, ] # same as above
```

Check the 'sex' variables:

```{r}
table(z$sex.x, z$sex.y)
```

The entire set of variables:

```{r}
names(z)
```

Tossing out the duplicate variables:

```{r}
all_cols <- names(z)
z <- z[, !(all_cols %in% c("age.y", "sex.y"))]
names(z)
```

Let's clean up the '.x' column names:

```{r}
gsub("\\.x", "", names(z))
names(z) <- gsub("\\.x", "", names(z))
names(z)
```

Save this file:

```{r}
write.csv(z, "atus_social.csv", row.names = FALSE)

# x <- read.csv("atus_social.csv", as.is = TRUE)
# names(x)
```

## Looking at Social_Time vs. Age

```{r}
plot(social_time ~ age, data=z)
```

We can make the points "transparent" to better
visualize denser areas:

```{r}
plot(social_time ~ age, data=z, col=rgb(0,0,0, 0.2), pch=16)
```

With a simple linear regression model, we might be
fitting the following equation:

$$\widehat{\text{social_time}} = \hat\beta_0 + \hat\beta_1 \text{age}$$


In R, the coefficients can be estimated by using the `lm()` function.

```{r}
lm(social_time ~ age, data=z)
```

According to R, the equation can be written as:

$$\widehat{\text{social_time}} = 130.80 + 3.55 \text{age}$$

```{r}
plot(social_time ~ age, data=z, col=rgb(0,0,0, 0.2), pch=16)
abline(a=130.80, b=3.55, lwd=2, col="red")
```

What about some other lines?

```{r}
plot(social_time ~ age, data=z, col=rgb(0,0,0, 0.2), pch=16)
abline(a=131, b=3.7, lwd=2, col="blue")
abline(a=200, b=0, lwd=2, col="green")
```

Residual is observed y minus predicted y.

$$e_i = y_i - \hat y_i$$

More precisely,

$$e_i = y_i - (\hat \beta_0 + \hat \beta_1 x_i)$$

For **least squares**, we want to minimize the sum of the squared residuals. We pick $\hat \beta_0$ and $\hat \beta_1$ such that $RSS=\sum e_i^2$ is minimized. 

$$RSS(\beta_0, \beta_1) = \sum_i (y_i - ( \beta_0 +  \beta_1 x_i))^2$$

We choose $\hat\beta_0$ and $\hat \beta_1$ to minimize the above quantity.

Assumptions for inference using linear regression.
LINE assumptions:

* Linearity - do the variables appear to be linearly associated?

* Independence - are the observations independent

* Normality

Usually we check a histogram of the residuals:

```{r}
m1 <- lm(social_time ~ age, data=z)
summary(m1)
hist(m1$resid)
```

Residuals, mathematically, will sum to 0. 

* Equal variance

Residuals should appear to have about constant variance throughout. This below is a plot of $e$ against $\hat y$.

```{r}
plot(resid(m1) ~ fitted(m1), col=rgb(0,0,0,0.2),
     pch=16)
```

Not a great model. For example, this plot is showing that we would be under-predicting for lower values of $\hat y$. 

Try a squared term for age:

```{r}
m2 <- lm(social_time ~ age + I(age^2), data=z)
m2

x <- data.frame(age = seq(0, 85, by=1))
head(x)
yhats <- predict(m2, x)
head(yhats)

plot(social_time ~ age, data=z, col=rgb(0,0,0,0.2))
lines(x$age, yhats, lwd=2, col="red")
```

