---
title: "NYC Flight Data"
author: "STAT 230"
date: "3/30/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r}
library(nycflights13)

library(dplyr)
library(tidyr)
library(ggplot2)
```

## The Data

```{r}
dim(flights)
flights
summary(flights)
```

### Filtering by Row

`filter()` allows us to filter a dataset by row.

Suppose we want to just look at flights that departed on New Year's Day.

```{r}
newyears <- filter(flights, month == 1, day == 1)

newyears <- filter(flights, month == 1 & day == 1)
head(newyears)
tail(newyears)

newyears <- flights %>% filter(month == 1, day == 1)
```


### Selecting columns

`select()` allows us to extract one or more columns from a data frame.

```{r}
newyears %>% select(dep_delay)

colnames(newyears)
newyears %>% select(starts_with("arr"))
newyears %>% select(contains("arr_"))

newyears %>% select(7:9)
newyears %>% select(arr_time:arr_delay)
newyears %>% select(-(arr_time:arr_delay))

newyears %>% select(-starts_with("arr"))
```

Getting distinct values of `carrier`:

```{r}
newyears %>% select(carrier) %>% distinct()
```


### Arranging the Data

```{r}
newyears %>% arrange(sched_dep_time, dep_delay)
newyears %>% arrange(sched_dep_time, desc(dep_delay))

```




### Renaming Columns


```{r}
names(newyears)

newyears %>% rename(hr = hour, min = minute)
```



### Adding new columns

```{r}
newyears <- newyears %>% mutate(gain = arr_delay - dep_delay)

# the old way of doing this:

# newyears$gain <- newyears$arr_delay - newyears$dep_delay
```


### Summarizing Data

`summarize()` can be used to compute summary statistics based on one or more columns in the data frame.

```{r}
flights %>% summarize(delay = mean(dep_delay, na.rm=TRUE),
                      numdests = n_distinct(dest))
```

By hour... what are these summary statistics?

```{r}
flights %>% group_by(hour) %>% 
  summarize(delay = mean(dep_delay, na.rm=TRUE),
                      numdests = n_distinct(dest))
```

Be careful when inserting line breaks. It's best to break after the piping operator and not before.

## Some Practice

Q: How many flights go out of NYC each day?


```{r}
daily <- flights %>% group_by(month, day) %>% 
  summarize(numflights = n())

library(lubridate)
flights <- flights %>% mutate(date = ymd(paste(year, month, day)))
head(flights$date)
tail(flights$date)

daily <- flights %>% group_by(date) %>% 
  summarize(numflights = n())
ggplot(daily, aes(x=date, y= numflights)) + geom_point()
```



How does this break down by day of the week?

```{r}
ggplot(daily, aes(x=date, y= numflights, 
                  color=wday(date, label=TRUE))) +
  geom_point()

```



Q: Suppose I am flying out of NYC on New Year's Day. Which airline carriers had the longest average arrival delay?

```{r}
newyears %>% group_by(carrier) %>% 
  summarize(meandelay = mean(arr_delay, na.rm=TRUE))
```






(This is not a well-posed question... why?)

Let's just consider flights to FLL (Fort Lauderdale):

```{r}
newyears %>% filter(dest == "FLL") %>% 
  group_by(carrier) %>% 
  summarize(meandelay = mean(arr_delay, na.rm=TRUE))
```

Or a plot:

```{r}
newyears %>% filter(dest == "FLL") %>%
  ggplot(aes(x= carrier, y = arr_delay)) +
  geom_boxplot()
```




Q: What do the NA's mean?

```{r}
newyears %>% filter(is.na(arr_delay)) %>% select(-(1:3))
```

It appears that some NAs are legitimate cancellations. Others might be fixable (with some work), but we'll choose to remove the NAs at this point.

```{r}
flights <- flights %>% filter(!is.na(arr_delay) & !is.na(dep_delay))
dim(flights)
```






Q: Consider the flights between NYC and BOS (Boston) each day. What are the average delays per hour of the day for each carrier?

```{r}
tmp <- flights %>% filter(dest == "BOS") %>% group_by(carrier, hour) %>%
  summarize(meandelay = mean(arr_delay))

ggplot(tmp, aes(x=hour, y=meandelay, color=carrier)) + geom_point()
```

Break this down further by airport of origin:

```{r}
tmp <- flights %>% filter(dest == "BOS") %>% 
  group_by(carrier, hour, origin) %>%
  summarize(meandelay = mean(arr_delay))

ggplot(tmp, aes(x=hour, y=meandelay, color=carrier)) + geom_point() + facet_wrap(~origin)
```


Break this down further by day of the week:

```{r}
tmp <- flights %>% filter(dest == "BOS") %>% 
  mutate(day_of_week = wday(date, label = TRUE)) %>%
  group_by(carrier, hour, origin, day_of_week) %>%
  summarize(meandelay = mean(arr_delay))

ggplot(tmp, aes(x=hour, y=meandelay, color=carrier)) + geom_point(alpha=0.5) + facet_grid(origin ~ day_of_week) +
  theme(legend.position = "none")
```


Q: How many flights are available at each of these combinations of variables?

```{r}
tmp <- flights %>% filter(dest == "BOS") %>% 
  mutate(day_of_week = wday(date, label = TRUE)) %>%
  group_by(carrier, hour, origin, day_of_week) %>%
  summarize(nflights = n())

ggplot(tmp, aes(x=hour)) + geom_bar(aes(weight=nflights, fill = carrier)) + facet_grid(origin ~ day_of_week) +
  theme(legend.position = "none")
```
