---
title: "Linear mixed-effects models"
author: "STAT 230"
date: "April 20, 2017"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE)
```

This is a brief introduction to a pretty big subject.
To find more you can do Google searches like

  "mixed effects models" r (nlme OR lme4) 

E.g. you can find anything from a nice gentle tutorial here:

  https://arxiv.org/ftp/arxiv/papers/1308/1308.5499.pdf

to the full story in the book "Mixed-Effects Models in S and S-PLUS" by Pinheiro and Bates.


# Simulating an example: male and female blood pressures

## Generating a data set

```{r}
set.seed(230)
nm <- nf <- 10
n <- nm+nf
timesm <- sample(1:10, nm, replace=TRUE)
timesf <- sample(1:10, nf, replace=TRUE)
times <- c(timesm, timesf)
id <- rep(1:n, times=times)
idfac <- as.factor(id)
N <- length(id)
male <- rep(c(1,0), times=c(sum(timesm), sum(timesf)))
beta0 <- c(125, 0)
# muf=beta0[1]; mum=beta0[1]+beta0[2]
sigb0 <- 12
b <- rnorm(n, 0, sigb0)
bb <- b[id]
sige0 <- 10
e <- rnorm(N, 0, sige0)
bp0 <- beta0[1] + beta0[2]*male + bb
bp <- bp0 + e
```

Data with the truth (we would not see the bp0 column):
```{r}
dat0 <- data.frame(id, idfac, male, bp0, bp) 
```

Data as we would see it:
```{r}
dat <- data.frame(id, idfac, male, bp)
```

Start a plot:
```{r}
jit <- .15
plot(id + runif(N,-jit,jit), bp, xlab='id')
for(j in 1:n){
  abline(v=j, col=gray(.8))
}
points(1:n, unique(bp0), pch="*", cex=3)
abline(v=nm+0.5, col='red')
```

## Some inappropriate analyses






## Show the sample means for each person





## Mixed effects models





## Looking further at what we got

```{r}
# BLUPs (fitted values from lme model)
s1 <- summary(m1)
names(s1)
m1$fitted
m1$fitted[,'id']
points(id, m1$fitted[,'id'], pch=19, col='blue')

# show the fitted fixed effects
m1$coef$fixed  # or:
fixef(m1)
lines(c(1,nm),rep(sum(fixef(m1)),2), col='blue')
lines(c(nm+1,n),rep(fixef(m1)[1],2), col='blue')
# So the predicted random effects are shrunk toward the estimated group means.

# the 20 BLUPs:
pred = unique(m1$fitted[,'id'])
# or:
pred = fixef(m1)[1] + 
  fixef(m1)[2] * rep(c(1,0),c(nm,nf)) + 
  ranef(m1)[[1]]
bps.true = beta0[1] + (beta0[2] * ((1:n)<=nm)) + b

fixed.est <- fixef(m1)[1] + fixef(m1)[2] * rep(c(1,0),c(nm,nf))

cbind(bps.true, fixed.est, sample.means, pred)
shrink <- (pred - fixed.est)/(sample.means - fixed.est)
ind=1:n
tmp <- data.frame(ind, times, shrink)
tmp[order(times,ind),]
```

How do the BLUPs do compared to the sample means?
```{r}
mean((sample.means-bps.true)^2)
mean((pred-bps.true)^2)
```

## A simulation: repeat the above, say 200 times, and record relevant results

```{r}
set.seed(230)
nit <- 200
results <- matrix(NA, nit, 4)
colnames(results) <- c("pvalbad", "pval", "sample_means_MSE", "lme_preds_MSE")
for(it in 1:nit){ 
  nm <- nf <- 10
  n <- nm+nf
  timesm <- sample(1:10, nm, replace=TRUE)
  timesf <- sample(1:10, nf, replace=TRUE)
  times <- c(timesm, timesf)
  id <- rep(1:n, times=times)
  idfac <- as.factor(id)
  N <- length(id)
  male <- rep(c(1,0), times=c(sum(timesm), sum(timesf)))
  beta0 <- c(125, 0)
  sigb0 <- 12
  b <- rnorm(n, 0, sigb0)
  bb <- b[id]
  sige0 <- 10
  e <- rnorm(N, 0, sige0)
  bp0 <- beta0[1] + beta0[2]*male + bb
  bp <- bp0 + e
  dat <- data.frame(id, idfac, male, bp)
  dat <- group_by(dat, id)
  sample.means <- summarize(dat, mean=mean(bp))$mean
  mbad <- lm(bp ~ male)
  smbad <- summary(mbad)
  pvalbad <- smbad$coefficients["male",4]
  m1 <- lme(bp ~ male, random=~1|id, data=dat)
  sm1 <- summary(m1)
  pval <- sm1$tTable["male", "p-value"]
  # the 20 BLUPs:
  pred = fixef(m1)[1] + 
    fixef(m1)[2] * rep(c(1,0),c(nm,nf)) + 
    ranef(m1)[[1]]
  bps.true = beta0[1] + beta0[2] * ((1:n)<=nm) + b
  MSE_sample_means <- mean((sample.means-bps.true)^2)
  MSE_lme_preds <- mean((pred-bps.true)^2)
  results[it,] <- c(pvalbad, pval, MSE_sample_means, MSE_lme_preds)
}
```

Look at how we did:

```{r}
results <- data.frame(results)
hist(results$pvalbad)
hist(results$pval)
sort(results$pvalbad)[1:20]
sort(results$pval)[1:20]
hist(results$lme_preds_MSE/results$sample_means_MSE)
abline(v=1, col=2, lwd=2)
hist(log2(results$lme_preds_MSE/results$sample_means_MSE))
abline(v=0, col=2, lwd=2)
```

# Science survey data

School Science Survey Data on attitudes to science, from a survey containing results for students from classes in private schools and in public schools.
From DAAG package.

  * PrivPub: a factor with levels private school, public school
  * school: a factor, coded to identify the school
  * class: a factor, coded to identify the class
  * sex: a factor with levels f, m
  * like: a summary score based on two of the questions, on a scale from 1 (dislike) to 12 (like)
  * Class: a factor with levels corresponding to each class

Get the data:
```{r}
science <- read.csv("http://www.stat.yale.edu/~jtc5/STAT230/data/science_DAAG.csv")
str(science)
for(j in c(2,3,6)){
  science[,j] <- as.factor(science[,j])
}
str(science)

sum(is.na(science))
s <- na.omit(science)
```

How many public and private schools?
```{r}
table(s$school, s$pp)
table(s$school, s$pp) > 0
colSums(table(s$school, s$pp) > 0)
```



